---
title: "Stat 199 Final Project"
author: "Team - Marley Dooling, Carter Harrigian, Mario Alverez"
output: pdf_document
---
```{r load-packages, message=FALSE}
library(tidyverse)
library(dplyr)
library(infer)
```

***Dataset 1- Social Media Usage of Politicians***
```{r data-1, message= FALSE}
Social_media_statistics_for_members_of_the_116th_Congress <- 
  read_csv("data/appendix csvs/Social media statistics for members 
           of the 116th Congress.csv")
Top_100_terms_associated_with_engagement_boosts_in_each_party <- 
  read_csv("data/appendix csvs/Top 100 terms associated with engagement 
           boosts in each party.csv")
```
**Data**: This data is sourced from the Pew Research Center, and it details 
every Facebook post and tweet published by all official and unofficial accounts 
maintained by every member of the U.S. Senate and House of Representatives 
from Jan. 1, 2015, to May 31, 2020. This data set contains almost 1.5 million 
Facebook posts, 3.3 million tweets, and additional data from 1,388 and 1,362 
congressional Facebook and Twitter accounts, respectively. For our research 
question, the relevant variables to consider are political party, social media 
platform, both "percent effect" variables, and term (referring to words used on
social media, not length of term in congress). 

**Research Question**: Is there a correlation between political party, social
media platform usage, specific term usage, and engagement (as measured by 
"percent effect" on reactions/favorites and shares/retweets)? Null hypothesis 
would be that there is no statistically significant correlation between party 
affiliation and social media activity, while the alternative would be that there 
is a correlation between the variables.

***Dataset 2- UN Labor Force Participation***
```{r dataset-2, message = FALSE}
laborgender <- read.csv('data/laborgender.csv')%>%
  select(-X)
```
**Data**: This data is sourced from the United Nations datamart, and includes
values only from the year 2015 and only for countries where data was able to be
collected, and therefore excludes many failed African and Asian states, among
others. It is a by-country and by-region analysis of various metrics relating
to women's civil rights and health, as well as their labor participation rate.

**Research Question**: What are the social, economic, physical, and political
factors most strongly impacting the labor force participation rate of women
worldwide? Null hypothesis would be that these factors do not affect labor 
force participation, and alternative hypothesis is that they do have an affect 
on the participation in labor force.

***Dataset 3- NBA***
```{r dataset-3, message = FALSE}
nba_raptor_play <- read.csv('data/nba-raptor/modern_RAPTOR_by_player.csv')
```
**Data**: This data comes from the RAPTOR prediction model (Robust Algorithm
using Player Tracking and On/off Ratings) that aims to predict player success 
through modern NBA metrics of plus-minus, win-shares, and wins above replacement
. The newest metric is on/off data, which analyzes team performance while 
player is on and off the floor, and it was designed to help understand player 
value for general managers and teams looking to trade. Height, age, and 
position are not elements of this dataset, so it just examines raw impact 
of players on the court with a higher weight on box score performance (score).

**Research Question**: What RAPTOR elements are most crucial for team success 
in recent years, and how do defensive and offensive success according to RAPTOR 
translate into team and player success? The null hypothesis would be that 
neither defense nor offense is more important to modern NBA win-shares / success
, and the alternative hypothesis is that either defense or offense is more 
relevant to modern NBA success.


