---
title: "The Predictors of Equitable Female Labor Force Participation"
author: "Lucky Charms - Marley Dooling, Carter Harrigian, Mario Alvarez"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
# The Determinants of Female Labor Force Participation Rate Parity

```{r load-packages, message=FALSE, echo = FALSE}
library(tidyverse)
library(dplyr)
library(infer)
library(patchwork)
library(broom)
library(fivethirtyeight)
library(car)
library(olsrr)
library(ggpubr)
```

```{r importset, echo = FALSE}
laborgender <- read.csv('data/laborgender.csv')%>%
  select(-X)

laborgender <- laborgender[, c(1,2,3,4,5,6,7,8,15,9,10,11,12,13,14,16,17,18,19)]
```

```{r scaledset, echo = FALSE, warning = FALSE}
laborgenderscaled <- laborgender%>%
  mutate(GII = GII*100,
         HDI = HDI*100,
         EconomyFrag = EconomyFrag*10,
         EconIneq = EconIneq*10,
         PublicServFrag = PublicServFrag*10,
         HumanRightsFrag = HumanRightsFrag*10,
         Fertility = Fertility*10,
         SecondaryRatio = SecondaryRatio*100,
         LogDiff = if_else(LBFPRDiff > 0, log(LBFPRDiff), 0))%>%
  filter(Country != "Saint Lucia" & Country != "Tonga")%>%
  select(-TertiaryRatio)
```

# Introduction and data

## Question and context

Our focal research question is **"What social, political,**
**physical, and economic factors affect the labor force participation rate of**
**women around the world and its parity to male participation?"**

Our approach to arriving at the answer will be multifaceted: 

First, we will recognize the limitations of our data-set and extrapolate data
based on entries we know to be missing and our knowledge of geopolitics. We will 
also mention how our model fits the best it possibly could.

Secondly, we will create appropriate visuals to understand and explain the
distribution of factors and relationships between them. 

Thirdly, the crux of our research will focus on labor force participation rates, specifically evaluating how they correlate to other factors like teen pregnancy, 
female representation in politics, etc. 

By using the quantitative methods covered in introductory data science course 
(among them visualizations and linear models), our research seeks to analyze
the complex relationship between gender dynamics and economic trends. Since the 
full story and factors that surround gender inequality will often be unique to 
culture and region, our data and conclusions will have some overarching 
limitations. It is always difficult to draw conclusions on a global variable, 
and we believe that the models we created are best fitting even with the idea 
of cultural differences looming as a downside of the results.

Our **hypothesis** at the onset are that lower measures
of gender equality (unequal), political stability, and human development will 
lead to lower female labor force participation rates and higher disparities 
between males and females.

## Data Source and Components

From the United Nations Datamart[1], we constructed a dataset 
combining factors such as labor force participation rates, fertility rates, 
measures of development, and others. Data collected is either from 2015 or 
from an interval including it: Fertility rates in particular are for the 
2010-2015 period.

Each row describes a country, its continent, and 16 other numeric variables
ranging from indicators of political fragility to measures of development.

The most crucial variables at the onset are:

**HDI**: Human Development Index. From 0 to 1, measures a country's level of
  development chiefly by lifespan, education level, and gross national income
  per-capita.
  
**LBFPRFem**: Female labor force participation rate.

**LBFPRDiff**: How much higher, in percentage points, the male labor force
  participation rate is than the female rate.
  
**FSI**: The Fragile States Index. Higher values indicates a country is more
  vulnerable to conflict and governmental collapse.[3]

All variables are explained, with units, on the accompanying README.

# Methodology

## Constraints and Methods

We begin our exploration by putting forth that all statistical analysis will be
conducted at the $\alpha$ = 0.10 significance level, and confidence intervals
will be at the 90% confidence level. These levels were picked with the variable
and unpredictable nature of human society in mind, and this allows us to still 
draw strong conclusions with variables that will have lots of variability as 
global data does.

The early visualizations have two primary goals: to understand the scope of 
global data and to allow the reader to see the general trends before any 
intervals or modeling is done. These graphs and plots are meant to be the 
reasoning behind and segway into more complex models that are meant to 
extrapolate data to draw more meaningful conclusions. The strength of the 
predictors that we aim to showcase through visualizations are evidence in 
support of the model that we create. With this said, we will attempt to maximize
strength of the model while maintaining diversity type of statistics calculated.

We will focus on analyzing **LBFPRDiff**, a variable indicating the rate
disparity between female and male labor force participation in an attempt to 
see which social, political, physical, and economic measures drive disparity.

## Establishing a Baseline

Faceting Data by region gives good context as to the true variability in global 
data and how regionally dependent similar data points tend to be. This high 
variability also explains the lower r-squared value that we will see later on. 

**Figure 1**

```{r Regions, echo = FALSE, fig.height = 4, fig.width = 8}
laborgender %>% 
  ggplot(mapping = aes(x = LBFPRDiff)) +
    geom_histogram(binwidth = 3, color = "pink") + 
    facet_wrap(~ Continent) +
    labs(x = "Participation Rate Disparity", 
         y = "Count", 
       title = "Distribution of Disparity in Female Labor Force Participation", 
       subtitle = "Percentage By Region") +
    theme_light()
```

We can also do the same thing by looking at women in government (positions
of power) to get a better handle on Women in the Labor Force. 
We will again group by region to see if there are any trends. 

**Table 0A**

```{r labormean, echo = FALSE}
labormean <- laborgender %>%
  group_by(Continent) %>%
  mutate(meancont = mean(WomenInGovt)) %>%
  select(Continent, meancont) %>%
  slice(1) 
labormean
```

```{r labormeanbar, echo = FALSE, fig.width = 8, fig.height = 3}
ch1 <- labormean %>%
  ggplot(mapping = aes(x = Continent, y = meancont)) + 
  geom_col(fill = "grey", color = "darkblue") + 
  theme_minimal() +
  labs(title = "Percentage of Women in Gov't by Continent", subtitle = "
       Using National values to Compute Average", x = "Continent", 
       y = "Percentage of Women in Gov't")
```

**Table 0B**

```{r HDImean, echo = FALSE}
HDImean <- laborgender %>%
  group_by(Continent) %>%
  mutate(meanhdi = mean(HDI)) %>%
  select(Continent, meanhdi) %>%
  slice(1) 
HDImean
```

```{r HDImeanplot, echo = FALSE}
ch2 <- HDImean %>%
  ggplot(mapping = aes(x = Continent, y = meanhdi)) + 
  geom_col(fill = "grey", color = "darkblue") + 
  theme_minimal() +
  labs(title = "Mean Human Development Index", subtitle = "
       Using National values to Compute Average", x = "Continent", 
       y = "HDI Avg")
```

By visualizing women in power and HDI values throughout continents, we can see 
if the bar graphs look similar. This would insinuate that these statistics go 
hand-in-hand as past research suggests.

**Figure 2**

```{r Means, echo = FALSE, message = FALSE, warning = FALSE, fig.height= 4, fig.width = 10}
(ch1) + (ch2)
```

To analyze the trends between the disparity between female and male labor
force participation rates and other variables, we first address a hypothesis
and visualize the possible correlation between the rate difference and human
development within our data.

**Note: Rate is calculated (Male rate - Female rate)** 

```{r HDIeffect, echo = FALSE}
HDIdiff <- laborgender%>%
  ggplot(mapping = aes(x = HDI, y = LBFPRDiff))+
  geom_point()+
  geom_smooth(se = FALSE, method = lm)+
  labs(title = "HDI vs. Rate disparity in labor force participation",
       x = "HDI", y = "Rate difference")
```

```{r FSIeffect, echo = FALSE}
FSIdiff <- laborgender%>%
  ggplot(mapping = aes(x = FSI, y = LBFPRDiff))+
  geom_point()+
  geom_smooth(se = FALSE, method = lm)+
  labs(title = "FSI vs. Rate disparity in labor force participation",
       x = "FSI", y = "Rate difference")
```

**Figure 3**

```{r HDIFSI, echo = FALSE, message = FALSE, warning = FALSE, fig.height= 4, fig.width = 10}
(HDIdiff) + (FSIdiff)
```

The basic scatterplot shows a slightly negative relationship between HDI and
the Rate disparity, implying that on average a higher HDI lowers disparity,
and also an opposing slightly positive relationship between FSI and the Rate
disparity.

Looking at the plot, there is a notable increased variance in Rate disparity
both when the HDI is low and when the FSI is high, implying that less developed
and more unstable countries have another range of factors affecting them
that merit deeper analysis for our entire data set.

```{r SDs, echo = FALSE, message = FALSE, warning = FALSE}
sd1 <- laborgender%>%
  filter(HDI <= 0.6)%>%
  summarize(sd = sd(LBFPRDiff))
sd2 <- laborgender%>%
  filter(HDI > 0.6)%>%
  summarize(sd = sd(LBFPRDiff))
sd1$sd - sd2$sd
```

The standard deviation in the interval HDI $\le$ 0.6 is 6.32 units higher than
the interval above it. The need to justify this leads into the linear model
section of our research, in which we attempt to find which appropriate
variables cause the most variability in the rate disparity.

To accomplish this, we will use the same dataset but with a scale on each
variable allowing coefficients to be interpreted more easily. We first remove
some irrelevant variables that have either a perfect or unquantifiable
relationship with rate disparity.

**Table 1**

```{r DiffReg, echo = FALSE}
DiffReg <- lm(LBFPRDiff ~ FSI + WomenInGovt + Fertility,
              data = laborgenderscaled)
glance(DiffReg)
tidy(DiffReg, conf.level = 0.9, conf.int = TRUE)
```

An r-squared value above 0.10 is generally one that accepted within 
the social sciences sphere [4] because of the variability we discussed earlier.
However in search of a better model, we shed FSI and HDI as possible values for
a linear regression model and analyze other continuous data to investigate 
the original hypothesis.


```{r augmenter_diff, echo = FALSE}
DiffReg_aug <- augment(DiffReg)
```

```{r independence_check, echo = FALSE}
c1 <- ggplot(DiffReg_aug, aes(x = seq(nrow(DiffReg_aug)), y = .resid)) + 
  geom_point() + 
  labs(title = "Independence Check", x = "Index", y = "Residual value") +
  theme_minimal()
```

```{r equal_variance_and_linearity_check, echo = FALSE}
c2 <- ggplot(DiffReg_aug, mapping = aes(x = .fitted, y = .resid)) +
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", lty = 2) +
  labs(title = "Equal Variance and Linearity Check", x = "Predicted Value", y = "Residual value") +
  theme_minimal()
```

```{r normality_check_1, echo = FALSE}
c3 <- ggplot(DiffReg_aug, mapping = aes(x = .resid)) +
  geom_histogram(binwidth = 2, fill = "grey", color = "darkblue") + 
  labs(title = "Normality Check (Histogram)", x = "Residuals", y = "Count") +
  theme_minimal()
```

```{r normality_check_2, echo = FALSE}
c4 <- ggplot(DiffReg_aug, mapping = aes(sample = .resid)) +
  stat_qq() + 
  stat_qq_line() +
  labs(title= "Normality Check (Linear Regression)")+
  theme_minimal()
```

**Figure 4**

```{r modelchecks, fig.height = 4, fig.width = 8, echo = FALSE}
(c1 + c2) / (c3 + c4)
```

After conducting these tests, we arrive at a multiple regression linear model that
holds the rate disparity as a relative function of the fragility of human
rights infrastructure, the teen birth rate, and the percentage of women in
the legislature. However, we notice above that the residuals have a distinct
right skew. After testing each variable to determine which predictor was responsible
for this skew, we determined that all were somewhat right skewed. This is 
most likely because some of the more developed nations are skewing the data 
more right.

```{r predictorcheck, echo = FALSE}
DiffReg_1 <- lm(formula = LBFPRDiff ~ (0*HumanRightsFrag) + (0*WomenInGovt) + TeenBirthRate, 
    data = laborgenderscaled)
DiffReg_1_aug <- augment(DiffReg_1)

DiffReg_2 <- lm(formula = LBFPRDiff ~ (HumanRightsFrag) + (0*WomenInGovt) + (0*TeenBirthRate), 
    data = laborgenderscaled)
DiffReg_2_aug <- augment(DiffReg_2)

DiffReg_3 <- lm(formula = LBFPRDiff ~ (0*HumanRightsFrag) + WomenInGovt + (0*TeenBirthRate), 
    data = laborgenderscaled)
DiffReg_3_aug <- augment(DiffReg_3)
```

```{r TBRCheck, echo = FALSE}
d1 <- ggplot(DiffReg_1_aug, mapping = aes(x = .resid)) +
  geom_histogram(binwidth = 2, fill = "grey", color = "darkblue") + 
  labs(title= "Normality Checks", subtitle = "Teen Birth Rate", x = "Residuals", y = "Count") +
  theme_minimal()
```

```{r HRFCheck, echo = FALSE}
d2 <- ggplot(DiffReg_2_aug, mapping = aes(x = .resid)) +
  geom_histogram(binwidth = 2, fill = "grey", color = "darkblue") + 
  labs(subtitle = "Human Rights Fragility", x = "Residuals", y = "Count") +
  theme_minimal()
```

```{r WIGCheck, echo = FALSE}
d3 <- ggplot(DiffReg_3_aug, mapping = aes(x = .resid)) +
  geom_histogram(binwidth = 2, fill = "grey", color = "darkblue") + 
  labs(subtitle = "Women in Government", x = "Residuals", y = "Count") +
  theme_minimal()
```

**Figure 5**

```{r varchecks, fig.height = 4, fig.width = 8, echo = FALSE}
(d1 + d2)/(d3)
```
As these visualizations illustrate, our model needed to be altered to better
fit our data. Therefore, we adjusted our model such that it holds the natural 
logarithm of the rate disparity as a regression function of other variables, 
using software to achieve the highest possible adjusted $R^2$. This also 
would help with the rightward skew that we see on the q-q plot.

**Table 2**

```{r optimalmodel, echo = FALSE}
DiffReg2 <- lm(LogDiff ~ . -LBFPRFem -Country -Continent -LBFPRMal - LBFPRTotal -GII -HDI -LBFPRDiff,
               data = laborgenderscaled)
DiffReg2 <- ols_step_forward_p(DiffReg2, penter = 0.1)$model
tidy(DiffReg2, conf.int = TRUE, conf.level = 0.9)
glance(DiffReg2)
```
Then, we ran regression diagnostics again using our updated model. As the graphs
below illustrate, this adjusted model better fits our data, and will thus allow
us to make more accurate predictions about labor force participation disparities.

```{r varcheck_DiffReg2, echo = FALSE}
DiffReg2_aug <- augment(DiffReg2)

e1 <- ggplot(DiffReg2_aug, aes(x = seq(nrow(DiffReg2_aug)), y = .resid)) + 
  geom_point() + 
  labs(title = "Independence Check", x = "Index", y = "Residual value") +
  theme_minimal()+
  ylim(-4,4)

e2 <- ggplot(DiffReg2_aug, mapping = aes(x = .fitted, y = .resid)) +
  geom_point() + 
  geom_hline(yintercept = 0, color = "red", lty = 2) +
  labs(title = "Equal Variance and Linearity Check", x = "Predicted Value", y = "Residual value") +
  theme_minimal()+
  ylim(-4,4)

e3 <- ggplot(DiffReg2_aug, mapping = aes(x = .resid)) +
  geom_histogram(binwidth = 0.25, fill = "grey", color = "darkblue") + 
  labs(title = "Normality Check (Histogram)", x = "Residuals", y = "Count") +
  theme_minimal()

e4 <- ggplot(DiffReg2_aug, mapping = aes(sample = .resid)) +
  stat_qq() + 
  stat_qq_line() +
  labs(title= "Normality Check (Linear Regression)")+
  theme_minimal()+
  ylim(-4,4)
```

**Figure 6**

```{r fig.height = 4, fig.width = 8, echo = FALSE}
(e1 + e2) / (e3 + e4)
```

The sole remaining weakness in our model are a handful of very low residuals
that exist as a result of logarithm properties in analyzing countries with
very low disparities. We opt to leave them rather than further adjust our
model so as to not misrepresent any other data. Also, the q-q plot 
fits better as was mentioned prior.

**Figure 7**

```{r augmentdiff, echo = FALSE, fig.height = 4.5, fig.width = 8}
tibble (
  Fertility = median(laborgenderscaled$Fertility),
  FSI = laborgenderscaled$FSI,
  WomenInGovt = laborgenderscaled$WomenInGovt
)%>%
  augment(DiffReg2, newdata = .)%>%
  ggplot(mapping = aes(x = WomenInGovt, y = FSI, color = .fitted))+
  geom_point()+
  scale_color_gradient(low = "darkgreen", high = "red")+
  labs(title = "Rate disparity as a function of stability and political representation", x = "Proportion of women in legislature",
       subtitle = "Holding fertility constant at its median",
       y = "Fragile States Index",
       color = "Predicted log of rate disparity")
```
Holding fertility constant at its median, we visualize here the negative
relation between women's representation and the rate disparity, and the positive
relationship between FSI and rate disparity, both in logarithmic terms. We
remove **Fertility** from the plot due to limitations on multidimensional
visualization while noting that it, too, has a negative effect on the log
of the rate disparity.

# Results

## Analyzing Disparities

**Table 3**

```{r tidier, echo = FALSE}
tidy(DiffReg2, conf.int = TRUE, conf.level = 0.9)
```

**Table 4**

```{r OddsLog, echo = FALSE}
tidy(DiffReg2) %>%
  mutate(estimate_mult = exp(estimate)) %>%
  select(term, estimate, estimate_mult, starts_with("conf"))
```


Firstly, we analyze and write out the results of our final log-based model. We
see that, **on average**, and holding the rest of variables constant: 

An increase of one tenth in **FSI** increases the log-rate
disparity by 0.0113, with a 90% confidence interval ranging from 0.00393 to 
0.0186. The disparity is therefore multiplied by around 1.011 for the 
same increase.

An increase of one percentage point in **women's representation in legislature**
decreases the log-rate disparity by 0.0254, with a 90% confidence interval
ranging from 0.0146 to 0.0362. The disparity is therefore multiplied 
by around 0.975 for the same increase.

An increase of **one tenth of a birth per woman** decreases the log-rate
disparity by 0.0275, with a 90% confidence interval ranging from 0.0153 to
0.0398. The disparity is are therefore multiplied by around 0.973 for the 
same increase, or around 0.760 for a **one-birth** increase.

We write these predictors out, returning to the original scale:

$$LogDiff = 3.331 + 0.0113\times{FSI} - 0.0254\times{WomenInGovt} - 0.275\times{Fertility}$$

Also, the latter measurements were derived from returning to a linear rate
disparity variable by exponentiation of both sides of the equation, like so:

$$LBFPRDiff = e^{3.331 + 0.0113\times{FSI} - 0.0254\times{WomenInGovt} - 0.275\times{Fertility}}$$

Where, by exponent rules, we can separate each term of the right-hand exponent
and propose that corresponding increases/decreases in the variable result in
multiplications of the rate disparity.

We reiterate per our statistical methods that all our predictors have p-values
of under 0.1, **women's representation** having 0.000167, Fertility having
0.000325, and FSI having 0.0124.

By our adjusted $R^2$, we conclude that our model accounts for 22.9% of the
variability in the log-rate disparity, a value that is not only the highest of
our displayed models but also the highest possible, as any removal or addition
would reduce this value. We are therefore satisfied with it, particularly given
our topic[4]. We discuss the possibility of improvement in further research
in the next section.

Now returning with this model to analyze on a global scale, we recall from our
methodology and dataset analysis that the more undeveloped regions of Asia,
Africa and Oceania suffer from the least equitable representations in
government, as well as the highest measures of instability and a tendency
toward higher fertility rates. Then, per our model, these regions will have
the highest rate disparities, as the mitigating effect of fertility is not
sufficient to counterbalance their substantial rises in instability and
decreases in female representation.

## Discussion

From our research, we were able to ascertain that increases in a country's
political stability and safety, female representation in government, and
fertility are conducive to lower disparities in labor force participation. 
Therefore, a country with a stable internal system where women 
own a representative share of governance will enjoy the lowest disparities, with
another mitigating effect coming from fertility, indicating that
perhaps countries where women birth more also expect women to work some
manner of job and contribute to their communities. Overall, stable countries
with higher standards of female self-determination will, in general, enjoy
the lowest disparities.

Despite our research's thoroughness, our data suffers from much the same
handicap that UN data does - many unstable or struggling countries do not
have appropriate data collection facilities for one or more of the variables
and therefore cannot always be included in considerations. Secondly, it is
unlikely that even 17 variables can strongly predict a trend in human and
societal behavior. Thirdly, while we analyzed labor force participation rates
and parity, we did not focus on quality of work or wages and their own parity.
Fourthly, we concede that our model remains appropriate but becomes more
erratic for less developed countries, indicating the need for more variables
to analyze.

With this reality in mind, we evaluated additional trends we noticed globally.
To begin, we notice a weak positive correlation between Fragile States Index (FSI)
and labor force participation rate disparities between men and women. This 
suggests that, in general, countries that are more vulnerable to conflict and 
governmental collapse have a workforce containing a lower proportion of women
(relative to men). This finding supports our hypothesis that lower
measures of gender equality, political stability, etc. lead to greater disparities
in labor force participation rates. Furthermore, we also found that "development"
-as measured by the Human Development Index (HDI)-is weakly negatively correlated
with rate disparities. Essentially, this means that as nations become increasingly
more developed, men and women are slightly more equally represented in the
workforce.
This finding further supports our hypothesis that increasing human development
will lower labor-force participation disparities between men and women.

We concede also that all variables in our dataset are intricately tied with
each other, and so achieving the intercept value is unlikely as, for example,
an FSI of 0 is theoretically impossible to achieve
alongside 0 women in government and a 0 teen birth rate. Therefore, our
intercept itself is not realistic and should be viewed more as a faux initial
value, and our variables should not be viewed in isolation from one another.

We opted not to include these interaction effects due to both difficulties
in visualization and unacceptably high variance inflations.

## Reflecting

If we were to be able to do this over again we would likely zone in 
on a particular region. Global data is difficult to manipulate because we 
don't want to exclude data points, but also want meaningful visualizations.
Being as specific as possible when using data in the humanities allows 
us to draw more meaningful conclusions, and if we were to do this project again 
we know that "meaningfulness" would be a primary goal of the results and 
visualizations. We would also use data that is less tied, but in this sphere 
it is hard to find data points that don't use similar metrics.

In summary, we found that an increase in political instability has an
exacerbating effect on the the disparity between female and male labor force
participation rates, while an increase in women's representation in government
and an increase in fertility both have mitigating effects.

We end this paper with a call for further research to use improved
visual software and more thorough and expansive data - which may take years
to be collectible - to better predict women's power in labor markets worldwide.

## References

- 1.) United Nations Datamart.
  https://data.un.org
  
  2.) Participation Rate vs. Unemployment Rate: What's the Difference?
  https://www.investopedia.com/ask/answers/061515/what-key-difference-between-participation-rate-and-unemployment-rate.asp
  
  3.) Fragile States Index 2015
  https://fundforpeace.org/2015/06/17/fragile-states-index-2015-the-book/
  
  4.) R- squared values within social sciences
  https://medium.com/@vince.shields913/why-we-dont-really-care-about-the-r-squared-in-econometrics-social-science-593e2db0391f